Ferromagnetism and spin excitations in topological Hubbard models with a flatband

### 2018-10-30 paper/paper.tex
arXiv version.

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in topological Hubbard models. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction below which the ferromagnetic phase is unstable, and obtain the ferromagnetic spin-1 excitation spectra of the Chern Hubbard model and the $Z_2$ Hubbard model. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, in both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. We elaborate that the destabilization of the ferromagnetic phase happens when the energy of the induced roton-like modes becomes zero with the increase of the nonflatness.
\end{abstract}
\maketitle

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998}. It was proved by Tasaki and Mielke that the ground state of a flat electron band with its filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards, this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state were also studied \cite{KA_PRL1994,SGDL_PRB2018} and itinerant topological magnons have been reported quite recently \cite{SGDL_PRB2018}.

\par The interplay between flatband ferromagnetism and nontrivial band topology enriches the related physics. In fact, ferromagnetism is essential in the generation of stable fractional Chern insulators in the proposals where the spin degrees of freedom of electrons are considered \cite{TMW_PRL2011,NSRCM_PRB2011,LBFL_PRL2012}. Furthermore, ferromagnetism can also lead to possible high-temperature quantum anomalous Hall effect (QAHE) when the nearly-flat topological band is half-filled \cite{NSRCM_PRL2012}. In this paper, we study the ferromagnetism and  spin excitations from the ferromagnetic ground state in nearly-flat topological bands. As a paradigm, we consider a square lattice Hubbard model whose free part is a $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). When the model is quarter filled (or correspondingly, the lower nearly-flat band is half filled), the ground state is spin fully polarized due to the ferromagnetism and exhibits QAHE because of the nonzero Chern number of a single-spin band. Consequently, the charge excitations are gapped and the low energy physics is dominated by the one-spin-flip excitations. These spin-1 excitations have been studied by a generalized bosonization scheme where the interacting fermionic model is mapped to a free bosonic model describing spin-wave excitations at the harmonic approximation \cite{DG_PRB2015}. The ferromagnetism was shown to be stable against such spin wave excitations, which are gapless in the Chern Hubbard model and gapped in the $Z_2$ Hubbard model. However in this bosonization scheme, the free part of the electron model plays no role in the spin wave excitations other than contributing a global constant, suggesting that it should fail due to the competition between the kinetic energy and potential energy of electrons when the nonflatness of the electron bands is not negligible \cite{T_PRL1994}. In fact, in a strictly local periodic tight-binding model, an energy band with a nonzero Chern number cannot be exactly flat \cite{CMST_JPMT2014}. Therefore it remains an open question on whether the ground state is stable against the spin-1 excitations and how the nonflatness of the electron bands manifests itself in the spin-1 excitation spectra in such models.

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical magnitude of the Hubbard interaction is found for both the Chern Hubbard model and the $Z_2$ Hubbard model, below which the ferromagnetic phase is unstable. Furthermore, the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which leads to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase. We also make a comparison of our results in the flatband limit with those obtained by the bosonization scheme \cite{DG_PRB2015}. Qualitative agreements are observed for the Chern Hubbard model, yet notable discrepancy appears for the $Z_2$ Hubbard model. We attribute the discrepancy to the self-interactions of the magnons, which are ignored in the bosonization scheme but treated exactly in our method after the projection.

\par The rest of the paper is organized as follows. In Sec. \ref{mm}, we introduce the Chern Hubbard model and $Z_2$ Hubbard model studied in this paper, discuss the nontrivial band topology of their free parts, interpret the emergence of QAHE resulting from the interplay between flatband ferromagnetism and nontrivial band topology, and formulate the exact diagonalization method with a projection onto the lower nearly-flat band on details. In Sec. \ref{nr}, we summarize the phase diagram and elaborate the spin-1 excitation spectra of both models. Section \ref{sd} provides a summary and discussion.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{lattice}
\caption{(Color online) (a) Schematic representation of the $\pi$-flux model. Blue and red solid circles denote the $A$ and $B$ sublattices, respectively. The nearest-neighbor hopping amplitudes (solid black lines) are equal to $t_{1}\exp(i\alpha_{\sigma}\pi/4)$ (see text) along the direction of the arrows, the next-nearest-neighbor hopping amplitudes are equal to $t_{2}$ (dashed green lines) and $-t_{2}$ (dashed purple lines). The shaded area denotes the unit cell. (b) First Brillouin Zone. $\Gamma=(0,0)$, $X_{1,2}=(\pm\pi,0)$, $Y_{1,2}=(0,\pm\pi)$, $M_{1,2,3,4}=(\pm\pi,\pm\pi)$, $O_{1,2,3,4}=(\pm\frac{\pi}{2},\pm\frac{\pi}{2})$.}
\label{lattice}
\end{figure}

\par We consider a generalized $\pi$-flux Hubbard model on the square lattice, whose Hamiltonian can be written as $H=H_0+H_U$, where $H_0$ is the spinfull genralization of the original spinless model proposed in Ref. \onlinecite{NSCM_PRL2011},
\begin{equation}
H_0=\sum_{\langle ij\rangle,\sigma}(t_1^{ij,\sigma}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.})
    +\sum_{\langle\langle ij\rangle\rangle,\sigma}(t_2^{ij}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.}),
\end{equation}
and $H_U$ is the Hubbard interaction
\begin{equation}
H_U=U\sum_{i}n_{i\uparrow}n_{i\downarrow}.
\end{equation}
Here, $c_{i\sigma}^{\dagger}(c_{i\sigma})$ creates (annihilates) a spin $\sigma$ electron at site $i$, $n_{i\sigma}=c^{\dag}_{i\sigma}c_{i\sigma}$ is the particle-number operator, $\langle ij\rangle$ denotes the nearest-neighbor (NN) bonds and $\langle\langle ij\rangle\rangle$ denotes the next-nearest-neighbor (NNN) bonds. As shown in Fig. \ref{lattice}(a), the spin-dependent NN hopping amplitude $t_1^{ij,\sigma}$ and the spin-independent NNN hopping amplitude $t_2^{ij}$ are given by
\begin{equation}
t_1^{ij,\sigma}=t_1\exp\left(i\delta_1^{ij}\alpha_\sigma\pi/4\right)
\end{equation}
and
\begin{equation}
t_2^{ij}=t_2\delta_2^{ij},
\end{equation}
respectively. Here, $\delta_1^{ij}=+1$ if the NN electron hopping is along the direction of the solid black arrow and $\delta_1^{ij}=-1$ if along the reversed direction. $\delta_2^{ij}=+1$ if the NNN electron hopping is along the dashed green lines and $\delta_2^{ij}=-1$ if along the dashed purple lines. The spin-dependent phase factor $\alpha_\sigma$ breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=\alpha_\downarrow=+1$, whereas it preserves the time-reversal symmetry but breaks the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=+1$ and $\alpha_\downarrow=-1$.

@@

\subsection{Chern Hubbard model}\label{CHM}
\begin{figure}
\includegraphics[width=\columnwidth]{fmcispectrum}
\caption{(Color online) (a) Spin-1 excitation spectra of the Chern Hubbard model in the FM phase. Green solid lines denote the spin waves and the grey region denotes the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Blue solid lines denote the upper and lower boundaries of the Stoner continuum determined by the band-splitting picture shown in Fig. \ref{bspicture}(a). Black arrows mark the local minima of the Stoner continuum. (b$_1$-b$_4$) Spectral weights for the lowest four eigen levels of the spin-1 excitation spectra with $\mathbf{q}=(0,0)$. The parameters are $t_1=1.0$, $t_2=0.490$ and $U=2.3$.}
\label{fmcispectrum}
\end{figure}

\par In this subsection, we focus on the quarter-filled Chern Hubbard model. In Fig. \ref{fmcispectrum}(a), we present the spin-1 excitation spectra in the FM phase along a high symmetry path in the first Brillouin zone with the parameters marked by the red star in Fig. \ref{phase}(a). We find that there are two kinds of obviously different excitations: one contains two low-lying modes labeled by the green solid lines exhibiting well defined band structures,  and the other forms a high-energy continuum as labeled by the shaded area. In fact, for these two kinds of modes, the patterns of the contributions from each spin-1 particle-hole scattering channel of electrons, which are embodied in the eigenvectors $\Psi_{\mathbf{q}}(\mathbf{k}_i)$ of Eq. (\ref{PHP}), are quite different. In Fig. \ref{fmcispectrum}(b), the spectral weights of each scattering channel, i.e. $|\Psi_{\mathbf{q}}(\mathbf{k}_i)|^2$ as a function of $\mathbf{k}_i$, for the four lowest levels with the center-of-mass momentum $\mathbf{q}=(0,0)$ ($\Gamma$ point) are shown. It is clear that for the modes in the green solid lines [Fig. \ref{fmcispectrum}($\text{b}_1$) and Fig. \ref{fmcispectrum}($\text{b}_2$)], the spectral weights come from a quite broad range of scattering channels; indeed, the spectral weight for the lowest mode at $\Gamma$ point is even homogeneous for all scattering channels. However, for those in the shaded area [Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4$)], the spectral weights almost come from a single scattering channel. Therefore, the modes in the green solid lines are collective spin-1 excitations and are identified as the spin waves, while the modes in the shaded area are individual spin-1 excitations and are identified as the Stoner continuum.

\par For comparison, the corresponding spin-1 excitation spectra in the flatband limit are also shown in Fig. \ref{fmcispectrum}(a) as the green dashed lines, which consist of low-lying spin waves and high-energy Stoner continuum as well. For both the original case and the flatband-limit case, the spin waves consist of an acoustic branch and an optical branch due to the $AB$ sublattice of the model. The acoustic branch is gapless at the $\Gamma$ point and disperses quadratically away from but near the $\Gamma$ point, which is the character of a ferromagnetic excitation with a Goldstone mode as the spin-fully polarized ground state spontaneously breaks the spin $SU(2)$ rotation symmetry. Additionally, the acoustic branch and the optical branch are degenerate along the $X_1$-$M_1$ path for both cases.

\par We also want to compare our results with that obtained by the generalized bosonization method at the harmonic approximation [see Fig. 4 in Ref. \cite{DG_PRB2015}]. Their result can be regarded as dispersions of free magnons, as in their scheme both the nonflatness of the $d$ band and the self-interactions between magnons are ignored. Apparently, their result is qualitatively consistent with ours in the flatband limit. However, considerable distinct features arise between theirs as well as ours in the flatband limit and ours in the dispersive case: excitation energies of the spin waves around the $X_1$ and $O_1$ points are strongly suppressed and new roton-like spin-wave excitations emerge. Furthermore, the boundaries of the Stoner continuum for the original model is dispersive and the local minima of its lower boundary coincides with the spin waves with suppressed energies, as indicated by the black arrows in Fig. \ref{fmcispectrum}(a). Note that in the flatband limit, only the nonflatness of the $d$ band is ignored, but the self-interactions between the magnons are treated exactly. Therefore, the new emerged roton-like spin waves must be attributed to the nonflatness of the $d$ band.

@@

\par The above physics can be illustrated by a simple band-splitting picture as shown in Fig. \ref{bspicture}(a). Roughly speaking, in the presence of ferromagnetism, the spin-down electron band and spin-up electron band are split in energy proportional to $U\langle m \rangle$, where $U$ is the Hubbard interaction strength and $\langle m \rangle$ is the average magnetic moment per site. For a quarter-filled electron model with two inequivalent sublattices, $\langle m \rangle=1/2$. Thus the energy splitting is $U/2$. Then an individual spin-1 excitation, or a mode in the Stoner continuum, can be viewed as the excitation of an electron from the lower fully-filled spin-up band to the upper empty spin-down band. This argument can be verified by our numerical results. In fact, the $\delta$ function-like distribution of the spectral weights of a mode in the Stoner continuum [as has be seen in Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4)$] implies that the excitation energy of such a mode can be approximated by the corresponding diagonal term in Eq. (\ref{PHP}). By numerics, we also find that the term $\frac{U}{N} \sum_{a=1,2} \sum_{\mathbf{p}\neq\mathbf{k}_i} \left| \mu_{a,\mathbf{p}\uparrow} \right|^2 \left| \mu_{a,\mathbf{k}_{i}-\mathbf{q}\downarrow} \right|^2 \sim \frac{U}{2}$. Therefore, the excitation energies of the Stoner continuum are roughly $\varepsilon_{d\downarrow} (\mathbf{k}_i-\mathbf{q}) + \frac{U}{2} -\varepsilon_{d\uparrow} (\mathbf{k}_i)$, which is consistent with the above picture. In Fig. \ref{fmcispectrum}(a), the lower and upper boundaries of the Stoner continuum approximated by this argument are plotted as the blue solid lines, which fits extremely well with the true boundaries of the shaded area, thus verifies the validity of this simple argument. Now we can find out the dominant scattering channels that contribute most to the local boundary bottoms of the Stoner continuum and read off their center-of-mass momentums easily. In Fig. \ref{bspicture}(b), we mark the local maxima of the spin-up band with red solid circles and the local minima of the spin-down band with purple ones. The scattering channels corresponding to the local bottoms marked by the black arrows in Fig. \ref{fmcispectrum}(a) are also indicated by green arrows. Obviously, a spin-1 excitation corresponding to the particle-hole scattering between a spin-up maximum and a spin-down minimum appears as a local boundary bottom of the Stoner continuum and the momentum difference of these two electron state provides the center-of-mass momentum of that spin-1 excitation. To understand the renormalization of the collective modes observed here, in Fig. \ref{bspicture}(c), we plot the spectral weights of the four lowest levels subtracted by those in the flatband limit with the center-of-mass momentum $q=(\pi,0)$ ($X_1$ point). The differences concentrate around the scattering channels labeled by $\mathbf{k}_i=(\pm\pi,0)$ and $\mathbf{k}_i=(0,\pm\pi)$. However, the results for the collective modes [Fig. \ref{bspicture}($\text{c}_1$) and Fig. \ref{bspicture}($\text{c}_2$)] are positive while those for the individual modes at the Stoner continuum boundary bottom [Fig. \ref{bspicture}($\text{c}_3$) and Fig. \ref{bspicture}($\text{c}_4$)] are negative. Therefore a noticeable amount of spectral weights transfer from the latter to the former, which indicates a strong coupling between the spin waves and the nearby Stoner continuum. This coupling renormalizes the energies of the spin waves and leads to the occurrence of a roton-like local minimum for the optical branch.

\begin{figure}
\includegraphics[width=\columnwidth]{pbcispectrum}
\caption{(Color online) (a$_1$) and (b$_1$): Spin-1 excitation spectra of the Chern Hubbard model near the FM/NFM phase boundary. Black arrow marks the minimum of the Stoner continuum. (a$_2$) and (b$_2$): Low-energy parts of the spin-1 excitation spectra shown in (a$_1$) and (b$_1$) along $\Gamma$-$X_1$ path with an amplified resolution. (a$_3$) and (b$_3$): Illustration of the splitting of the corresponding lower electron bands. $t_2=0.437$ for (a) and $t_2=0.430$ for (b). Other parameters are fixed at $t_1=1.0$, $U=2.3$.}
\label{pbcispectrum}
\end{figure}

>>>>

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in interacting topological insulators. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux state with topologically nontrivial and nearly-flat electron bands, which can realize either the Chern or $Z_2$ Hubbard model. By using the numerical exact diagonalization method with a projection onto the nearly-flat band, we obtain the ferromagnetic spin-1 excitation spectra for both the Chern and $Z_2$ Hubbard models, consisting of spin waves and Stoner continuum. The spectra exhibit quite distinct dispersions for both cases, in particular the spin wave is gapless for the Chern Hubbard model, while gapped for the $Z_2$ Hubbard model. Remarkably, in both cases, the nonflatness of the free electron bands introduces dips in the lower boundary of the Stoner continuum. It significantly renormalizes the energies of the spin waves around these dips downward and leads to roton-like spin excitations. We elaborate that it is the softening of the roton-like modes that destabilizes the ferromagnetic phase, and determine the parameter region where the ferromagnetic phase is stable.
\end{abstract}
\maketitle

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998}. It was proved by Tasaki and Mielke that the ground state of a flat electron band with a filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards, this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state were also studied \cite{KA_PRL1994,SGDL_PRB2018} and itinerant topological magnons have been reported quite recently \cite{SGDL_PRB2018}.

\par The interplay between flatband ferromagnetism and nontrivial band topology enriches the related physics. In fact, ferromagnetism is essential in the generation of stable fractional Chern insulators in the proposals where the spin degrees of freedom of electrons are considered \cite{TMW_PRL2011,NSRCM_PRB2011,LBFL_PRL2012}. Furthermore, ferromagnetism can also lead to possible high-temperature quantum anomalous Hall effect (QAHE) when the nearly-flat topological band is half-filled \cite{NSRCM_PRL2012}. In this paper, we study the ferromagnetism and  spin excitations from the ferromagnetic ground state in nearly-flat topological bands. As a paradigm, we consider a square lattice Hubbard model whose free part is a $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. Depending on the nearest-neighbor hopping, the model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). When the model is quarter filled (or correspondingly, the lower nearly-flat band is half filled), the ground state is spin fully polarized due to the ferromagnetism and exhibits QAHE because of the nonzero Chern number of a single-spin band. Consequently, the charge excitations are gapped and the low energy physics is dominated by the one-spin-flip excitations. These spin-1 excitations have been studied by a generalized bosonization scheme where the interacting fermionic model is mapped to a free bosonic model describing spin-wave excitations at the harmonic approximation \cite{DG_PRB2015}. The ferromagnetism was shown to be stable against such spin wave excitations, which are gapless in the Chern Hubbard model and gapped in the $Z_2$ Hubbard model. However in this bosonization scheme, the free part of the electron model plays no role in the spin wave excitations other than contributing a global constant, suggesting that it should fail due to the competition between the kinetic energy and potential energy of electrons when the nonflatness of the electron bands is not negligible \cite{T_PRL1994}. In fact, in a strictly local periodic tight-binding model, an energy band with a nonzero Chern number cannot be exactly flat \cite{CMST_JPMT2014}. Therefore it remains an open question on whether the ground state is stable against the spin-1 excitations and how the nonflatness of the electron bands manifests itself in the spin-1 excitation spectra in such models.

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical magnitude of the Hubbard interaction is found for both the Chern Hubbard model and $Z_2$ Hubbard model, below which the ferromagnetic phase is unstable. Furthermore, the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which leads to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase. Therefore, we elaborate the mechanism of the instability of this flat-band ferromagnetism as the softening of the emergent roton-like modes with the increase of noflatness. We also make a comparison of our results in the flatband limit with those obtained by the bosonization scheme \cite{DG_PRB2015}. Qualitative agreements are observed for the Chern Hubbard model, yet notable discrepancy appears for the $Z_2$ Hubbard model. We attribute the discrepancy to the multimagnon processes, which are ignored in the bosonization scheme but treated exactly in our method after the projection.

\par The rest of the paper is organized as follows. In Sec. \ref{mm}, we introduce the Chern Hubbard model and $Z_2$ Hubbard model studied in this paper, discuss the nontrivial band topology of their free parts, interpret the emergence of QAHE resulting from the interplay between flatband ferromagnetism and nontrivial band topology, and formulate the exact diagonalization method with a projection onto the lower nearly-flat band on details. In Sec. \ref{nr}, we discuss the phase diagram and elaborate the spin-1 excitation spectra of both models. Section \ref{sd} provides a summary and discussion.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{lattice}
\caption{(Color online) (a) Schematic representation of the $\pi$-flux model. Blue and red solid circles denote the $A$ and $B$ sublattices, respectively. The nearest-neighbor hopping amplitudes (solid black lines) are equal to $t_{1}\exp(i\alpha_{\sigma}\pi/4)$ (see text) along the direction of the arrows, the next-nearest-neighbor hopping amplitudes are equal to $t_{2}$ (dashed green lines) and $-t_{2}$ (dashed purple lines). The shaded area denotes the unit cell. (b) The first Brillouin zone. The next-nearest-neighbor distance is set to one, so $\Gamma=(0,0)$, $X=(\pi,0)$, $Y=(0,\pi)$, $M=(\pi,\pi)$, $O=(\frac{\pi}{2},\frac{\pi}{2})$.}
\label{lattice}
\end{figure}

\par We consider a generalized $\pi$-flux Hubbard model on the square lattice, whose Hamiltonian can be written as $H=H_0+H_U$, where $H_0$ is the spinfull genralization of the original spinless model proposed in Ref. \onlinecite{NSCM_PRL2011},
\begin{equation}
H_0=\sum_{\langle ij\rangle,\sigma}(t_1^{ij,\sigma}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.})
    +\sum_{\langle\langle ij\rangle\rangle,\sigma}(t_2^{ij}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.}),
\end{equation}
and $H_U$ is the Hubbard interaction
\begin{equation}
H_U=U\sum_{i}n_{i\uparrow}n_{i\downarrow}.
\end{equation}
Here, $c_{i\sigma}^{\dagger}(c_{i\sigma})$ creates (annihilates) a spin $\sigma$ electron at site $i$, $n_{i\sigma}=c^{\dag}_{i\sigma}c_{i\sigma}$ is the particle-number operator, $\langle ij\rangle$ denotes the nearest-neighbor (NN) bonds and $\langle\langle ij\rangle\rangle$ the next-nearest-neighbor (NNN) bonds. As shown in Fig. \ref{lattice}(a), the spin-dependent NN hopping amplitude $t_1^{ij,\sigma}$ and the spin-independent NNN hopping amplitude $t_2^{ij}$ are given by
\begin{equation}
t_1^{ij,\sigma}=t_1\exp\left(i\delta_1^{ij}\alpha_\sigma\pi/4\right)
\end{equation}
and
\begin{equation}
t_2^{ij}=t_2\delta_2^{ij},
\end{equation}
respectively. Here, $\delta_1^{ij}=+1$ if the NN electron hopping is along the direction of the solid black arrow and $\delta_1^{ij}=-1$ if along the reversed direction. $\delta_2^{ij}=+1$ if the NNN electron hopping is along the dashed green lines and $\delta_2^{ij}=-1$ if along the dashed purple lines. The spin-dependent phase factor $\alpha_\sigma$ breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=\alpha_\downarrow=+1$, whereas it preserves the time-reversal symmetry but breaks the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=+1$ and $\alpha_\downarrow=-1$.

@@

\subsection{Chern Hubbard model}\label{CHM}
\begin{figure}
\includegraphics[width=\columnwidth]{fmcispectrum}
\caption{(Color online) (a) Spin-1 excitation spectra of the Chern Hubbard model in the FM phase. Green solid lines denote the spin waves and the grey region denotes the Stoner continuum. Blue solid lines denote the upper and lower boundaries of the Stoner continuum determined by the band-splitting picture shown in Fig. \ref{bspicture}(a). Black arrows mark the local minima of the Stoner continuum. (b$_1$-b$_4$) Spectral weights for the lowest four eigen levels of the spin-1 excitation spectra with $\mathbf{q}=(0,0)$. The parameters are $t_1=1.0$, $t_2=0.490$ and $U=2.3$. Green dashed lines in (a) represent the corresponding spectra in the flatband limit.}
\label{fmcispectrum}
\end{figure}

\par In this subsection, we focus on the quarter-filled Chern Hubbard model. In Fig. \ref{fmcispectrum}(a), we present the spin-1 excitation spectra in the FM phase along a high symmetry path in the first Brillouin zone with the parameters marked by the red star in Fig. \ref{phase}(a). It is shown that there are two kinds of obviously different excitations: one contains two low-lying modes labeled by the green solid lines exhibiting well defined band structures,  and the other forms a high-energy continuum as labeled by the shaded area. We find that, for these two kinds of modes, the patterns of the contributions from each spin-1 particle-hole scattering channel of electrons, which are embodied in the eigenvectors $\Psi_{\mathbf{q}}(\mathbf{k}_i)$ of Eq. (\ref{PHP}), are quite different. In Fig. \ref{fmcispectrum}(b), the spectral weights of each scattering channel, i.e. $|\Psi_{\mathbf{q}}(\mathbf{k}_i)|^2$ as a function of $\mathbf{k}_i$, for the four lowest levels with the center-of-mass momentum $\mathbf{q}=(0,0)$ ($\Gamma$ point) are shown. It is clear that for the modes in the green solid lines [Fig. \ref{fmcispectrum}($\text{b}_1$) and Fig. \ref{fmcispectrum}($\text{b}_2$)], the spectral weights come from a quite broad range of scattering channels; indeed, the spectral weight for the lowest mode at $\Gamma$ point is even homogeneous for all scattering channels. However, for those in the shaded area [Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4$)], the spectral weights almost come from a single scattering channel. Therefore, the modes in the green solid lines are collective spin-1 excitations and are identified as the spin waves, while the modes in the shaded area are individual spin-1 excitations and are identified as the Stoner continuum. The spin waves consist of an acoustic branch and an optical branch due to the $AB$ sublattice of the model. The acoustic branch is gapless at the $\Gamma$ point and disperses quadratically away from the $\Gamma$ point, which is the character of a ferromagnetic excitation with a Goldstone mode as the spin-fully polarized ground state spontaneously breaks the spin $SU(2)$ rotation symmetry. Additionally, the acoustic and optical branches are degenerate along the $X$-$M$ path.

\par In the flatband limit, the spin-1 excitation spectra as shown in Fig. \ref{fmcispectrum}(a) as the green dashed lines show several similarities with the dispersive case, including the constitution of the spin waves and Stoner continuum exhibiting as a flat line at $E=0.5U$, the gapless acoustic spin wave and its quadratical dispersion  near the $\Gamma$ point, and the degeneracy between the acoustic and optical branches along the $X$-$M$ path. Comparing results in the flatband limit with those obtained by the generalized bosonization method at the harmonic approximation [see Fig. 4 in Ref. \cite{DG_PRB2015}], one will find that two results are qualitatively quite consistent. In the scheme adopted in Ref. \cite{DG_PRB2015}, the ignorance of the interactions between magnons suggests that only the single-magnon excitations are captured. So, this consistence indicates that the interactions between magnons shows a negligible effect in the flatband limit. However, when the free band becomes dispersive, considerable distinct features appears. Now, the spin-wave energies around the $X=(0,\pi)$ and $O=(\pi/2,\pi/2)$ points are strongly suppressed, as a result two local minima are formed at $X$ and $O$ which can be ascribed to the emergence of the roton-like spin-wave excitations. Furthermore, the continuum exhibiting as a straight line in the flatband limit extends to a large grey region shown in Fig. \ref{fmcispectrum}(a) with dispersive boundaries, and the local minima of its lower boundary coincide with those in the spin-wave dispersion, as indicated by the black arrows in Fig. \ref{fmcispectrum}(a).

@@

\par This observation shows that it is the extending of the Stoner continuum arising from the nonflatness of the $d$ band that pushes down the spin-wave excitations. In particular, it suggests that the couplings between the individual excitations and spin waves would be responsible for the emergence of the roton-like spin waves.

\par The formation of the dispersive boundary of the Stoner continuum can be understood by a simple band-splitting picture as shown in Fig. \ref{bspicture}(a). In the presence of ferromagnetism, the spin-down and spin-up electron bands are split in energy which is roughly proportional to $U\langle m \rangle$, with $U$ the Hubbard interaction and $\langle m \rangle$ the average magnetic moment per site. For a quarter-filled electron model with two inequivalent sublattices, $\langle m \rangle=1/2$. Consequently, an individual spin-1 excitation, i.e., a mode in the Stoner continuum, corresponds to the excitation of an electron from the fully-filled spin-up band to the empty spin-down band with an energy proportional to the difference in the initial and final states plus the band splitting $U/2$.
Indeed, the $\delta$ function-like distribution of the spectral weights of a mode in the Stoner continuum [as has be seen in Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4)$] implies that its excitation energy can be approximated by the corresponding diagonal term in Eq. (\ref{PHP}). Numerically, we also find that the term $\frac{U}{N} \sum_{a=1,2} \sum_{\mathbf{p}\neq\mathbf{k}_i} \left| \mu_{a,\mathbf{p}\uparrow} \right|^2 \left| \mu_{a,\mathbf{k}_{i}-\mathbf{q}\downarrow} \right|^2 \sim \frac{U}{2}$. Therefore, the excitation energies of the Stoner continuum are roughly $\varepsilon_{d\downarrow} (\mathbf{k}_i-\mathbf{q}) + \frac{U}{2} -\varepsilon_{d\uparrow} (\mathbf{k}_i)$, which is consistent with the above analysis. The lower and upper boundaries of the Stoner continuum determined in this way are plotted as the blue solid lines in Fig. \ref{fmcispectrum}(a), which fits extremely well with the true boundaries of the shaded area, thus verifies the validity of this simple argument.

\par With this picture, the position of a local boundary bottom of the Stoner continuum is determined by the transferred momentum of the particle-hole scattering between a spin-up maximum and a spin-down minimum. Thus, in Fig. \ref{bspicture}(b), we mark the local maxima of the spin-up band with red solid circles and the local minima of the spin-down band with purple ones. One can see that the transferred momenta coincide completely with the positions
 of the local bottoms in Fig. \ref{fmcispectrum}(a) which are indicated here by green arrows.

\par To understand the renormalization of the spin waves observed above, in Fig. \ref{bspicture}(c), we plot the spectral weights of the four lowest levels subtracted by those in the flatband limit at momentum $q=(\pi,0)$ ($X$ point). The differences concentrate around the scattering channels labeled by $\mathbf{k}_i=(\pm\pi,0)$ and $\mathbf{k}_i=(0,\pm\pi)$. However, the results for the collective modes [Fig. \ref{bspicture}($\text{c}_1$) and Fig. \ref{bspicture}($\text{c}_2$)] are positive while those for the individual modes at the Stoner continuum boundary bottom [Fig. \ref{bspicture}($\text{c}_3$) and Fig. \ref{bspicture}($\text{c}_4$)] are negative. Therefore a noticeable amount of spectral weights transfer from the latter to the former, which indicates a strong coupling between the spin waves and the nearby Stoner continuum. This coupling renormalizes the energies of the spin waves and leads to the occurrence of a roton-like local minimum for the optical branch.

\begin{figure}
\includegraphics[width=\columnwidth]{pbcispectrum}
\caption{(Color online) (a$_1$) and (b$_1$): Spin-1 excitation spectra of the Chern Hubbard model near the FM/NFM phase boundary. Black arrow marks the minimum of the Stoner continuum. (a$_2$) and (b$_2$): Low-energy parts of the spin-1 excitation spectra shown in (a$_1$) and (b$_1$) along the $\Gamma$-$X$ path with an amplified resolution. (a$_3$) and (b$_3$): Illustration of the splitting of the corresponding lower electron bands. $t_2=0.437$ for (a) and $t_2=0.430$ for (b). Other parameters are fixed at $t_1=1.0$, $U=2.3$.}
\label{pbcispectrum}
\end{figure}
